Towards Contract Modules for the Tezos Blockchain

### 2021-08-01 papers/execution-model/APLAS 2021/2021-aplas-rebuttal.txt

Thanks to the reviewers for their insightful comments.

In a revision, we will address all specific comments, except where otherwise noted.

## reviewer 58A

* We agree that mechanizing the typing-related results is an important step forward. We are currently looking into it, but it's ongoing and hence not reported in the paper.
* Our initial examples were related to Oracles, but the work has moved away from that and we will update the text and related work to reflect that.
* *off-chain script runs on behalf of a single entity* this is indeed the case in our implementation. We will emphasize this in the revised version.
* *current approaches to verifiable clients* excellent point! we will add this aspect to our related work.

*Is the fact that contracts are deterministic a strength or a weakness?* It's mostly a strength, except for applications that require nondeterminism or some degree of impredictability, the most prominent example being (pseudo) random numbers.

*implementation does not seem polished enough* It's in constant flux as we are working against a moving target. Tezos has been updating its protocol and its library infrastructure in quite rapid succession. We are happy to add the interface in an appendix if it fits in the page limit.

*implementation for chkBal/chkPuh/etc* These are metafunctions used to describe the semantics of the blockchain transactions. The model functionality that is implemented in the Tezos client or the node.

*present your work as an extension to Michelson and its corresponding formalisation* we think that the present work is an excellent starting point for this endeavor. This way we have a blueprint for the formalization.

## reviewer 58B

We are sorry that our text does not make the need for the operational semantics sufficiently clear. In the end, we wish to state and prove properties of a system comprising OCaml code (multi-threaded or distributed) connected  via the API to Michelson contracts on the Tezos blockchain. This work is a first step in obtaining insight into the problem, next will be a formalization that connects with the Mi-Cho-Coq formalization of Michelson contracts, as mentioned by reviewer 58A.

## reviewer 58C

Let us comment on the weaknesses.

1. We started with a mini language, but were forced to add features as our work proceeded. We'll see what we can cut down.
2. *customized decision procedure* I'm not sure where this impression comes from, as we have no customized decision procedure. We do not give any indication of that in the paper and there is nothing special going on in type checking beyond OCaml's provisions. It would be advantageous to have subtyping, but as OCaml does not support subtyping, we decided to keep types separate and rely on cast operations that are checked at run time. BTW this approach is inspired by Michelson's design, which has similar casting operations, e.g., `IMPLICIT_CONTRACT` which casts from type `address` to type `contract unit`.
3. Soundness. We have a sketch of the type soundness proof on paper, but nothing typeset so far. It is structured as usual in type preservation and progress results. Progress is, of course, relative to the behavior of the blockchain (which we model as a gray box), so that any operation that requires interaction with the blockchain may be stuck temporarily.
4. Implementation. We'll be happy to make some space to discuss implementation details.

### 2021-08-06 papers/contract-modules/fmbc2021/fmbc-main.tex
first round of revisions

%\ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}}
%%TODO mandatory: Please choose ACM 2012 classifications from
%%https://dl.acm.org/ccs/ccs_flat.cfm
\ccsdesc[100]{}

@@

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Contract modules provide a clean, language-integrated way to interact
with a blockchain. They abstract over underlying string-based
interfaces and details like fee handling. They provide a high-level
typed interface which reduces a contract invocation to a function call
in the language.

Contract modules do not provide a fixed API, but
rather generate a specialized interface for each contract. This
interface is statically checked against the contract implementation to
ensure type safety and exception safety (every failure condition
arising is handled by proper error reporting).

Our work is situated in the context of the Tezos blockchain, which
supports Michelson as its low-level contract language, and the
language OCaml, which comes with an expressive polymorphic type system
as well as a powerful module system that we
enhance with contract modules.

@@

Tezos is a third generation, account-based, self amendable
blockchain \cite{tezos-whitepaper}. It employs a proof-of-stake consensus protocol, which
includes ways to evolve the protocol itself. The consensus protocol is
executed by so-called bakers and their proposed blocks are checked by
validators. They receive some compensation in the form of
tokens (Tezzies) for their work. According to proof-of-stake,
bakers and validators are just nodes elected by the Tezos network
according to their token balance.

Each Tezos contract owns an account as well as some storage. Contracts
are pure functions of type parameter $\times$ storage $\to$ operation
list $\times$ storage. When a contract is invoked with a parameter,
the blockchain provides the current storage and updates it with the
second, storage component of its return value. The first component is
a list of blockchain operations (contract deployments, token
transfers, contract invocations) that are executed transactionally
after the first invocation terminates. Each invocation may be
accompanied with an amount of tokens that are added to the current
account balance of the contract.

Contracts are implemented in the language Michelson, a fully typed
stack-based language. Each contract has fixed types for its parameter
and for its storage. The storage is initialized when the contract is
deployed. Besides primitive types like unit, int, bool, address, and string,
there are pairs, sums, functions, lists, and maps (and many more) that
can serve as types for storage and parameters.

\section{An auction contract}
\label{sec:an-auction-contract}
\begin{lstlisting}[float,caption={Simple auction contract (auction.tz)},captionpos=b,label={lst:auction-header},language=michelson,numbers=none,emph={close,bid},emphstyle=\underbar]
parameter (or (unit %close) (unit %bid));
storage (pair bool          # bidding allowed
         (pair address      #  contract owner
          address           # highest bidder's address
        ));
\end{lstlisting}

@@

The contract works as follows. It is deployed with storage
\lstinline/(true,(owner,owner))/ which indicates that bidding is
allowed and the contract owner is currently also the highest
bidder. On deployment the owner deposits an initial balance to indicate the
minimum bid. Closing the contract transfers the balance to the
owner. It is restricted to the owner. Closing as well as bidding fails
if the auction is
closed. If bidding is open and the amount of tokens accompanying the bid exceeds the
current highest bid, the current bidder replaces the previous highest
bidder and the previous highest bidder is reimbursed. Otherwise,
bidding fails, too.

To invoke this contract from an OCaml program, we'd like to generate an
OCaml module, say \lstinline/Auction/, from a specification of the
contract. This module contains two functions
\lstinline/close/ and \lstinline/bid/ corresponding to the
entrypoints. The type of these entrypoints reflects further properties
of these entrypoints as well as the ways in which an entrypoint might
fail.

@@

We consider the technological failures like Java's unchecked
exceptions, but we wish to deal with the explicit failures like
checked exceptions~\cite{DBLP:conf/oopsla/AnconaLZ01}.
Our generated code handles failures in a suitable
error monad that makes the failures explicit in a custom
datatype.\footnote{Alternatively, this could be done using OCaml
  exceptions, but we chose to stay within the monadic framework that
  is already used by existing Tezos APIs.}

@@

  entrypoint close ()
  raises "closed"               (** auction closed *)
       | "not owner"            (** caller cannot close *)
end
\end{lstlisting}
Listing~\ref{lst:contract-module-example} shows a contract module for
the auction contract.  It declares the entrypoint \lstinline/bid/ as
\lstinline/paid/, i.e., it needs to be invoked with a
non-zero amount of tokens, it gives the pattern \lstinline/()/ for the
input value of type \lstinline/unit/, and it specifies two possible
failure messages that we wish to deal with programmatically.
The \lstinline/close/ entrypoint is similar, but requires no tokens.

@@

  val close
    : Tezos.pukh -> ?fee:Tezos.mutez
    -> (Tezos.status, close_errors) Tezos.monad
\end{lstlisting}
% end 
Listing~\ref{lst:generated-signature} contains an OCaml module
signature as it could be generated from the contract module. The
module \lstinline/Tezos/ supposedly contains types and other low-level
Tezos-specific definitions. The type \lstinline/pukh/ for public
key hashes identifies contracts, the type \lstinline/mutez/ stands for
Tezos tokens, the type \lstinline/status/ reflects the internal return
status, and \lstinline/monad/ is an internal monad type. The signature
declares a function and an error type for each entrypoint.

@@

We plan to check the contract by symbolic execution against its
specification in the contract module. Here are some examples of
checkable properties.

For each entrypoint, we collect the set of reachable instructions.
For example, the  \lstinline/AMOUNT/ instruction obtains the amount of
tokens sent with a contract invocation.
It should not be possible to reach that instruction from an unpaid
entrypoint  like \lstinline/close/.

@@

@@

As it is expensive to invoke a contract just to find out that it
fails, we propose to extend entrypoint specifications with
preconditions as shown in Listing~\ref{lst:safer-contract-module}. The
idea is that the generated OCaml module tries to check the
preconditions off-chain before invoking the contract. To this end, the
off-chain code needs to obtain properties like balance, storage etc of the
contract, but this information is available from the Tezos node without a fee!
We discuss two of the preconditions to highlight the properties that 
need to be analyzed.

@@

For the failure \lstinline/"to low"/, the analysis is very similar:
we need to know that there is no successful execution of
\lstinline/bid/ after an execution of \lstinline/close/. Moreover,
each invocation of \lstinline/bid/ raises the balance of the contract
monotonically. Thus, if the off-chain check
\lstinline/amount > balance/ fails, we can be sure that the contract
invocation will also fail; either because some closed the auction or
because the balance is at least as high as in the off-chain sample. 
\section{Related work}
\label{sec:related-work}
%RPC APIs, often provided by current blockchains such as Ethereum \cite{eth-whitepaper} and Tezos \cite{tezos-whitepaper}, use loosely structured data, such as a JSON-based format, which is difficult for a programmatic program to handle. As a result, there is increasing work on providing better programmatic interfaces to the blockchains. Web3.js \cite{web3.js} provides an Ethereum JavaScript API and offers Java Script users a convenient interface to interact with the Etherum blockchain. Later, Web3.py \cite{web3.py}, derived from Web3.js, is developed to provide a Python library for interacting with Ethereum. Tezos prove a PRC API that utilizes JSON format for both input and output data, which is neither convenient nor programmatic.

@@

Status: we are currently working on a prototype of the symbolic interpreter.
% As future work, we
% combine the contract module with our typed API designed to provide a
% general type-safe programmatic API for the Tezos blockchain.

>>>>

%%TODO mandatory: Please choose ACM 2012 classifications from
%%https://dl.acm.org/ccs/ccs_flat.cfm
\begin{CCSXML}
<ccs2012>
   <concept>
       <concept_id>10011007.10011006.10011060.10011690</concept_id>
       <concept_desc>Software and its engineering~Specification languages</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
   <concept>
       <concept_id>10011007.10011006.10011041.10010943</concept_id>
       <concept_desc>Software and its engineering~Interpreters</concept_desc>
       <concept_significance>300</concept_significance>
       </concept>
 </ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~Specification languages}
%\ccsdesc[300]{Software and its engineering~Interpreters}

@@

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{Bruno Bernardo and Diego Marmsoler}
\EventNoEds{2}
\EventLongTitle{3rd International Workshop on Formal Methods for Blockchains (FMBC 2021)}
\EventShortTitle{FMBC 2021}
\EventAcronym{FMBC}
\EventYear{2021}
\EventDate{July 18-19, 2021}
\EventLocation{Los Angeles, California, USA (Virtual Conference)}
\EventLogo{}
\SeriesVolume{95}
\ArticleNo{4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

We propose contract modules that provide a clean, language-integrated
way to interact with a blockchain from a host language (OCaml in our
case). They abstract over underlying string-based interfaces and
details like fee handling. They provide a high-level typed interface
which reduces a contract invocation to a function call in the host
language.

The contract modules approach does not provide a fixed API, but
rather generates a specific interface for each contract with one
function for each entrypoint of the contract. This
interface is statically checked against the contract implementation to
ensure type safety and exception safety. That is, values passed to an
interface function do not lead to type mismatches when invoking the
underlying contract. Morever, every failure condition arising during
contract execution is handled by proper error reporting according to
the interface.

Our work is situated in the context of the Tezos blockchain, which
supports Michelson as its low-level contract language, and the
host language OCaml, which comes with an expressive polymorphic type system
as well as a powerful module system that we
enhance with contract modules.

@@

Tezos is a third generation, account-based, self amendable
blockchain \cite{tezos-whitepaper}. It employs a proof-of-stake consensus protocol, which
includes ways to evolve the protocol itself. The consensus protocol is
executed by so-called bakers and their proposed blocks are checked by
validators. They receive some compensation in the form of
tokens (Tezzies) for their work. According to proof-of-stake,
bakers and validators are nodes elected by the Tezos network
according to their token balance.

Each Tezos contract is associated with an account as well as some
storage. Contracts are pure functions of type $\mathit{parameter} \times
\mathit{storage} \to \mathit{operation\ list} \times
\mathit{storage}$, where the types \textit{parameter} and
\textit{storage} are depend on the specific contract while the type
\textit{operation} is fixed by the Tezos system. When a contract is invoked with a parameter,
the blockchain provides the current storage and updates it with the
second, storage component of its return value. The first component of
the return value is a list of blockchain operations (contract deployments, token
transfers, contract invocations, and delegation of baking rights) that
are executed transactionally after the first invocation terminates. Each invocation may be
accompanied with an amount of tokens that are added to the current
account balance of the callee contract.

Contracts are implemented in the language Michelson, a statically typed
stack-based language. Each contract has fixed types for its parameter
and for its storage. The storage is initialized when the contract is
deployed. Besides primitive types like unit, int, bool, address, and string,
there are pairs, sums, functions, lists, and maps along with a range
of domain-specific types (operation, key, signature, timestamp,
key\_hash, contract, mutez --- for tokens, and so on) most of which
can serve as types for storage and parameters.

\section{An auction contract}
\label{sec:an-auction-contract}
\begin{lstlisting}[float,caption={Simple auction contract (auction.tz)},captionpos=b,label={lst:auction-header},language=michelson,numbers=none,emph={close,bid,bidding,owner,hi_bidder},emphstyle=\underbar]
parameter (or (unit %close) (unit %bid));
storage (pair (bool %bidding)
              (pair (address %owner)
                    (address %hi_bidder)));
\end{lstlisting}

@@

The contract works as follows. It is deployed with storage
\lstinline/(true,(owner,owner))/ which indicates that bidding is
allowed and the contract owner is currently the highest
bidder. On deployment the owner deposits an initial balance to indicate the
minimum bid. Closing the contract transfers the balance to the
owner. It is restricted to the owner. Closing as well as bidding fails
if the auction is
closed. If bidding is open and the amount of tokens accompanying the bid exceeds the
current highest bid, the current bidder replaces the previous highest
bidder and the previous highest bidder is reimbursed. Otherwise,
bidding fails, too.

To invoke this contract from an OCaml program, we generate an
OCaml module, say \lstinline/Auction/, from a specification of the
contract. This module contains two functions
\lstinline/close/ and \lstinline/bid/ corresponding to the
entrypoints. The type of these entrypoints reflects further properties
of these entrypoints as well as the ways in which an entrypoint might
fail.

@@

We consider the technological failures like Java's unchecked
exceptions, but we wish to deal with the explicit failures like
checked exceptions~\cite{DBLP:conf/oopsla/AnconaLZ01}.
Our generated code handles failures in a suitable
error monad that makes the failures explicit in a custom
datatype.\footnote{Alternatively, errors could be modeled using OCaml
  exceptions, but we choose to stay within the monadic framework that
  is used by existing Tezos APIs.}

@@

  entrypoint close ()
  raises "closed"               (** auction closed *)
       | "not owner"            (** caller cannot close *)
end
\end{lstlisting}
Listing~\ref{lst:contract-module-example} shows a contract module for
the auction contract.  It declares the entrypoint \lstinline/bid/ as
\lstinline/paid/, i.e., it should be invoked with a
non-zero amount of tokens, it states the pattern \lstinline/()/ for the
input value of type \lstinline/unit/, and it specifies two 
failure messages that we wish to deal with programmatically.
The \lstinline/close/ entrypoint is similar, but it is not
\lstinline/pair/. This contract reflects the understanding of the
programmer with the intention that the \lstinline/raises/ clauses
cover all failures that can arise during execution of the respective
entrypoint.

@@

  val close
    : Tezos.pukh -> ?fee:Tezos.mutez
    -> (Tezos.status, close_errors) Tezos.monad
\end{lstlisting}
% end 
Listing~\ref{lst:generated-signature} contains an OCaml module
signature as it would be generated from the contract module. The
module \lstinline/Tezos/ supposedly contains types and other low-level
Tezos-specific definitions. The type \lstinline/pukh/ for public
key hashes identifies contracts, the type \lstinline/mutez/ stands for
Tezos tokens, the type \lstinline/status/ reflects the internal return
status, and \lstinline/monad/ is an internal monad type. The signature
declares a function and an error type for each entrypoint.

@@

We check the contract by symbolic execution against its
specification in the contract module. Symbolic execution proceeds in
the usual way by calculating a symbolic stack at each transition from
one Michelson instruction to the next. The initial stack is calculated
from the storage type, the parameter type, and the entrypoint.
As we aim to keep symbolic
values as concrete as possible, some instructions may turn out to be
unreachable or only reachable under certain conditions. This way, we
obtain, for each entrypoint, a set of final symbolic stacks along with
a path condition indicating when this final state is
reachable. Moreover, for each \lstinline/FAILWITH/ instruction we
obtain a symbolic value for the reported message and a path condition
indicating reachability of this instruction. We employ the SMT solver
Z3 \cite{DBLP:conf/tacas/MouraB08,DBLP:conf/setss/BjornerMNW18} to
check the feasibility of a path condition.

Here are some simple examples of checkable properties.

For each entrypoint, we collect the set of reachable instructions.
For example, the  \lstinline/AMOUNT/ instruction obtains the amount of
tokens sent with a contract invocation.
It should not be possible to reach that instruction from an unpaid
entrypoint  like \lstinline/close/. This property is straightforward
to check from the path condition generated for the \lstinline/AMOUNT/
instruction.

@@

For the simple auction contract the symbolic interpreter flags an
omission in the contract module
(Listing~\ref{lst:contract-module-example}). The problem is that
bidding returns the previous highest bid as part of the transaction
where the highest bidder is stored as a value of type
\lstinline/address/.  However, to receive a token transfer, this
address must by cast (by the Michelson implementation of the contract)
to an implicit contract of Michelson type \lstinline/(contract unit)/. This cast is unavoidable as a
value of contract type cannot be stored. However, the cast may fail and its failure leads to an
error condition that is reported with a \lstinline/FAILWITH/ instruction that is not covered by the
contract module.\footnote{We generated the Michelson code for this example using the Liquidity
  compiler (\url{https://www.liquidity-lang.org/}). While the failures for \lstinline/closed/ and
  \lstinline/too low/ are explicit in the source program, the compiler inserts the casts into the contract
  type automatically using the error message \texttt{No entrypoint default with parameter type
    unit}.}  The entrypoint for closing has the same issue with the address of the owner, who is
scheduled to receive the balance of the contract.

The best way to address this issue would be to assert that the addresses stored for the owner and
the highest bidder always cast successly into the \lstinline/(contract unit)/ type so that the
stated module type can be retained. Unfortunately, doing so requires control over the deployment of
the contract where the storage is initialized and it is not clear whether deployment should be part
of the API generated from a contract module.

The quick fix is to add \lstinline/raises/ clauses for the \texttt{No entrypoing \dots}
message. With this fix, we successfully check the module description against the Michelson
implementation of its contract.

@@

As it is expensive to invoke a contract just to find out that it
fails, we propose to extend entrypoint specifications with
preconditions as shown in Listing~\ref{lst:safer-contract-module}. The
idea is that the generated OCaml module tries to check the
preconditions off-chain before invoking the contract. To this end, the
off-chain code needs to obtain properties like balance, storage etc of the
contract, but this information is available from a Tezos node without a fee! Of course, such an
off-chain check is prone to race conditions as concurrent contract invocations from other parties
may interfere and change the data before our module gets a chance to invoke the contract. 
We discuss two of the preconditions to highlight the properties that 
need to be analyzed and where race conditions may interfere.

@@

For the failure \lstinline/"to low"/, the analysis is very similar:
we need to know that there is no successful execution of
\lstinline/bid/ after an execution of \lstinline/close/. Moreover,
each invocation of \lstinline/bid/ raises the balance of the contract
monotonically. Thus, if the off-chain check
\lstinline/amount > balance/ fails, we can be sure that the contract
invocation will also fail; either because someone closed the auction or
because the balance is at least as high as in the off-chain sample. 
\section{Related work}
\label{sec:related-work}
%RPC APIs, often provided by current blockchains such as Ethereum \cite{eth-whitepaper} and Tezos \cite{tezos-whitepaper}, use loosely structured data, such as a JSON-based format, which is difficult for a programmatic program to handle. As a result, there is increasing work on providing better programmatic interfaces to the blockchains. Web3.js \cite{web3.js} provides an Ethereum JavaScript API and offers Java Script users a convenient interface to interact with the Etherum blockchain. Later, Web3.py \cite{web3.py}, derived from Web3.js, is developed to provide a Python library for interacting with Ethereum. Tezos prove a PRC API that utilizes JSON format for both input and output data, which is neither convenient nor programmatic.

@@

Status: the symbolic interpreter works on Michelson files which results from compiling the examples
provided with the Liquidity compiler. It generates human-readable output as well as output in the
SMTlib format suitable for SMT-solvers like Z3.

While our conceptual approch is applicable and would be beneficial for other blockchains, the actual
implementation is very much tied to the Tezos blockchain. The key asset here is the symbolic
interpreter which is hardcoded for Michelson and adapted to the peculiarities of the Tezos
blockchain. By targeting Michelson, our work is applicable to all languages running on Tezos, but
a similar tool would have to be developed from scratch for another blockchain.
